---
title: "Data wrangling 1"
author: "Claus O. Wilke"
date: last-modified
format: 
  revealjs:
    theme:
      - default
      - Wilke-theme.scss
    auto-stretch: false
    chalkboard: true
execute:
  fig-format: svg
  fig-asp: 0.618
  fig-width: 5.5
---

```{r}
#| include: false
library(tidyverse)
library(palmerpenguins)

# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(14))
update_geom_defaults("point", list(size = 2))
#update_geom_defaults("line", list(linewidth = 1))
```

## Elementary data manipulations

::: {.incremental}
- Pick rows: `filter()`
- Pick columns: `select()`
- Sort rows: `arrange()`
- Count things: `count()`
- Make new columns: `mutate()`
:::

# But first: the pipe operator `|>`

# `|>` is pronounced "and then"

## The pipe `|>` feeds data into functions

::: {.fragment .small-font}
```{r}
#| echo: true
library(palmerpenguins) # loads the `penguins` dataset

head(penguins)
```
:::

## The pipe `|>` feeds data into functions

::: {.small-font}
```{r}
#| echo: true
library(palmerpenguins) # loads the `penguins` dataset

# head(penguins)
penguins |>
  head()
```
:::

## The pipe `|>` feeds data into functions

::: {.small-font}
```{r}
#| label: penguins-scatter
#| echo: true
#| fig-cap: "&nbsp;"
#| out-width: "60%"
ggplot(penguins, aes(bill_length_mm, bill_depth_mm, color = species)) + 
  geom_point()
```
:::

## The pipe `|>` feeds data into functions

::: {.small-font}
```{r}
#| label: penguins-scatter2
#| echo: true
#| fig-cap: "&nbsp;"
#| out-width: "60%"
penguins |>
  ggplot(aes(bill_length_mm, bill_depth_mm, color = species)) +
  geom_point()
```
:::

## Old-school "magrittr" pipe: `%>%`

::: {.small-font}
```{r}
#| label: penguins-scatter3
#| echo: true
#| fig-cap: "&nbsp;"
#| out-width: "60%"
penguins %>%
  ggplot(aes(bill_length_mm, bill_depth_mm, color = species)) +
  geom_point()
```
:::


## Which to use? Native pipe or old-school pipe?

::: {.incremental}
- The native pipe `|>` is the future. I'll be using it throughout.
- The magrittr `%>%` works fine. You can use it if you llike to.
:::

<!-- Segment ends here -->

# Picking rows or columns, and sorting

## Pick rows from a table: `filter()`

::: {.move-down-2em}
![](../assets/data-wrangling-1/filter.svg){fig-align="center" width="75%"}
:::

## Filter out penguins of species Gentoo

::: {.small-font}
```{r}
#| echo: true
penguins |>
  filter(species == "Gentoo")
```
:::

## Filter out penguins with bill length > 50 mm

::: {.small-font}
```{r}
#| echo: true
penguins |>
  filter(bill_length_mm > 50)
```
:::

## Pick columns from a table: `select()`

::: {.move-down-2em}
![](../assets/data-wrangling-1/select.svg){fig-align="center" width="75%"}
:::


## Pick columns `species`, `island`, and `sex`

::: {.small-font}
```{r}
#| echo: true
penguins |>
  select(species, island, sex)
```
:::

## Sort the rows in a table: `arrange()`

::: {.move-down-2em}
![](../assets/data-wrangling-1/arrange.svg){fig-align="center" width="75%"}
:::

## Sort penguins by bill length, ascending

::: {.small-font}
```{r}
#| echo: true
penguins |>
  arrange(bill_length_mm)
```
:::

## Sort penguins by bill length, descending

::: {.small-font}
```{r}
#| echo: true
penguins |>
  arrange(desc(bill_length_mm))
```
:::

<!-- Segment ends here -->

# Counting things

## Count things

::: {.small-font}
```{r}
#| echo: true
penguins
```
:::

## Count things

::: {.small-font}
```{r}
#| echo: true
penguins |>
  count(species)
```
:::

## Count things

::: {.small-font}
```{r}
#| echo: true
penguins |>
  count(species, island)
```
:::

## Use the pipe to build analysis pipelines

::: {.small-font}
```{r}
#| echo: true
penguins |>
  filter(species == "Adelie")
```
:::

## Use the pipe to build analysis pipelines

::: {.small-font}
```{r}
#| echo: true
penguins |>
  filter(species == "Adelie") |>
  select(island, sex)
```
:::

## Use the pipe to build analysis pipelines

::: {.small-font}
```{r}
#| echo: true
penguins |>
  filter(species == "Adelie") |>
  select(island, sex) |>
  count(island, sex)
```
:::

<!-- Segment ends here -->

# Adding new columns to a table

## Make a new table column: `mutate()`

::: {.move-down-2em}
![](../assets/data-wrangling-1/mutate.svg){fig-align="center" width="75%"}
:::

## Example: flipper length in cm

::: {.small-font}
```{r}
#| echo: true
penguins |>
  select(species, island, flipper_length_mm)
```
:::

## Example: flipper length in cm

::: {.small-font}
```{r}
#| echo: true
penguins |>
  select(species, island, flipper_length_mm) |>
  mutate(flipper_length_cm = flipper_length_mm / 10)
```
:::

## Make multiple columns at once

::: {.small-font}
```{r}
#| echo: true
penguins |>
  select(species, island, flipper_length_mm) |>
  mutate(
    flipper_length_cm = flipper_length_mm / 10,  # <- notice the comma
    flipper_length_in = flipper_length_mm / 25.4
  )
```
:::

<!-- Segment ends here -->

## Further reading

- R for Data Science: [Chapter 5: Data transformation](https://r4ds.had.co.nz/transform.html)
- R for Data Science: [Chapter 18: Pipes](https://r4ds.had.co.nz/pipes.html)
- **dplyr** documentation: [Introduction to dplyr](https://dplyr.tidyverse.org/articles/dplyr.html)
- **dplyr** reference documentation: [One table verbs](https://dplyr.tidyverse.org/reference/index.html#section-one-table-verbs)
